NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-04-13 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex
edits to the counterion binding section
Update manuscriptPS.tex
Edits to the ca binding section

Calcium binding affinity to membranes containing the negatively charged PS lipids can be
experimentally measured by detecting the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the measurement is not complicated by the dehydrated lipid-ion complexes and phase separation and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of an ion-free reference state
in the presence of negatively charged lipids, our simulations give
coherent results for POPC headgroup order parameters as a function of
CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the
experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters
in POPC:POPS (5:1) mixtures as Ca$^{2+}$ concentration increases (Fig. \ref{changesWITHCaClPS}),
indicating overestimated calcium binding affinity.
The only exception is the CHARMM36 model with the NBfix
interaction employed for calcium~\cite{kim16}, which underestimates the modulation of order parameters, 
indicating weaker binding affinity than experiments.
%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity
is underestimated in CHARMM36 simulations when using the NBfix for calcium \cite{kim16}, but overestimated 
in all the other tested models. This is evident in the calcium density distributions where almost all Ca$^{2+}$ ions bind to the membrane interface in all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
the smaller $\alpha$-carbon. All these changes are significantly overestimated in the
tested simulation models, including CHARMM36 which underestimated the binding affinity.
Importantly, the different simulation models predict qualitatively different behaviour
for the POPS $\alpha$-carbon order parameters with added calcium.
For example, both order parameters decrease in Berger, but increase
in MacRog, and in Lipid14/17 and CHARMM36 a more complicated behavior is seen.
This is in contrast to the PC headgroup, where
qualitatively correct reponse to bound ions is observed
in all simulation models, despite significant discrepancies in the headgroup
structure without additional ions~\cite{catte16}. Therefore, we conclude that
improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.
\section{Conclusions}
Lipids with PS headgroups, and their interactions with ions,
play an important role in lipid-mediated signaling
processes \cite{leventis10,yeung08}. Attemps to use MD
simulations to interpret the spectroscopic data have produced contradictory results for the calcium binding details to PS
headgroups \cite{melcrova16,valentine18,hallock18}.  
Here (as was previously done for PC lipids \cite{botan15,catte16})
we used the headgroup C--H bond order parameters and the open collaboration approach to evaluate the quality
of the headgroup structure and ion binding affinity to PS lipids
in available MD force fields.
The main advantage of this approach is the direct connection
between the accurately measured experimental order parameters and the simulations,
which reduces the ambiguity in the interpretation of experiments.

>>>>

Calcium binding affinity to membranes containing the negatively charged PS lipids can be
experimentally quantified by measuring the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the measurement is not compromised by the dehydrated lipid-ion complexes and phase separation, and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of an ion-free reference state
in the presence of negatively charged lipids, our simulations give
coherent results for POPC headgroup order parameters as a function of
CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the
experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters upon increasing Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
indicating too strong calcium binding affinity.
The only exception is the CHARMM36 simulation utilizing the NBfix
interaction for calcium~\cite{kim16}, wherein the modulation of order parameters is underestimated, 
indicating weaker binding affinity than in experiments.
%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
Notably, CHARMM36 simulations with the NBfix corrections \cite{venable13,kim16} give similar binding affinities of
calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
is underestimated in CHARMM36 simulations when using the NBfix for calcium \cite{kim16}, but overestimated 
in all the other tested models. This is evident in the calcium density distributions where almost all Ca$^{2+}$ ions bind to the membrane interface in all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong dependence of CaCl$_2$ with small concentrations that saturates below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
tested simulation models, including CHARMM36 which underestimated the binding affinity.
Importantly, the different simulation models predict qualitatively different behaviour
for the two POPS $\alpha$-carbon order parameters with added calcium.
For example, both order parameters decrease in Berger, but increase
in MacRog, and in Lipid14/17 and CHARMM36 a more complicated behavior is seen.
This is in contrast to the PC headgroup, where
qualitatively correct reponse to bound ions is observed
in all simulation models, despite significant discrepancies in the headgroup
structures produced in salt-free simulations~\cite{catte16}. Therefore, we conclude that
improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.
\section{Conclusions}
Lipids with PS headgroups, and their interactions with ions,
play an important role in lipid-mediated signaling
processes \cite{leventis10,yeung08}. Attemps to use MD
simulations to interpret the spectroscopic data have produced contradictory results for the calcium binding details to PS
headgroups \cite{melcrova16,valentine18,hallock18}.  
Here (as was previously done for PC lipids \cite{botan15,catte16})
we used the headgroup C--H bond order parameters and the open collaboration approach to evaluate the quality
of the headgroup structure and ion binding affinity to PS lipids
in available MD force fields.
The main advantage of this approach is the direct connection
between the accurately measured experimental order parameters and the simulations,
which reduces the ambiguity in the interpretation of experiments.
